---
title: "Supplementary Material: R script for generating plots and overviews"
subtitle: "Mixtec Sound Change Database"
author: "Sandra Auderset"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document: 
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: no
    toc: false
mainfont: Linux Libertine
header-includes:
  - |
    ```{=latex}
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      breaksymbolleft={}, 
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
    }
    ```
csl: generic-style-rules-for-linguistics.csl
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy="styler",echo=TRUE,dev = "cairo_pdf",warning = FALSE,fig.align = 'center',out.width="0.9\\textwidth",out.height="\\textheight")
```

This script provides the code for generating the plots in the paper and aggregating the data for the use cases described there.
All maps are created with the package `ggmap' (Kahle & Wickham 2013) using Stamen maps from Stadia (https://stadiamaps.com/attribution/).

# Preparation

First, we load the necessary packages (cf. the Rmd file for details).
```{r preamble, include=FALSE}
# load packages
library(pacman)
p_load(cowplot,
       extrafont,
       formatR,
       ggmap, 
       ggrepel,
       ggtext,
       ggthemes,
       here,
       janitor, 
       kableExtra,
       stringi,
       styler, 
       tidybulk, 
       tidyverse,
       viridis,
       xtable)
```
```{r stadia, include=FALSE}
# API key for maps
register_stadiamaps("3a811dbc-3a12-4a65-a604-94e474a482af", write = FALSE)
```

We read in the files from the Mixtec Sound Change Database repository.
Next, we check the coverage of variables and languages and exclude those who have very low coverage from further analysis. Low coverage for varieties is defined here as having NA for more than a third of the variables. This is not currently the case in our database, so we can keep all the languages.
Low coverage for variables is defined here the same way, i.e. as having NA for more than one third of the varieties. For the current study, this leads to the exclusion of 29 variables (see details in code).
We create a file for analysis that excludes these low coverage variables.
```{r coverage}
# read in variable file
var_seg_all <- read_tsv("https://raw.githubusercontent.com/SAuderset/MixteCaSo/main/variables/variables_segments.tsv")

# summarize language coverage
total_number_changes <- ncol(select(var_seg_all, -DOCULECT))
# check for number of NA
coverage_lang <- var_seg_all %>%
  mutate(na_lang = rowSums(is.na(.))) %>%
  select(DOCULECT, na_lang) %>%
  arrange(desc(na_lang)) %>% 
  mutate(exclude = if_else(na_lang>(total_number_changes/3), "yes", "no"))
# none of the languages needs to be excluded

# summarize variable coverage
total_number_languages <- nrow(var_seg_all)
# exclude variables that cover less than 1/3rd of varieties
coverage_var <- var_seg_all %>%
  summarize(across(matches("[[:upper:]]{1}\\d{2}"), ~sum(is.na(.)))) %>%
  pivot_longer(everything())
# get names of low coverage variables; list them
low_coverage_var <- coverage_var %>%
  filter(value>(total_number_languages/3)) %>%
  pull(name)
low_coverage_var

# create data set for analysis excluding low coverage variables
var_seg <- var_seg_all %>%
  select(!c(low_coverage_var))
```

# Map of the sample with subgroup/dialect area membership and inset
We set up base map of the Mixtec region for plotting with the ggmap package (Kahle & Wickham 2013) with a stamen map.
We will exclude the diaspora variety Abosolo del Valle, located in Veracruz, from all maps, because including it would lead to less visibility for other varieties on the map. In the paper we will mention if it diverges from the rest of Group 71 (Mixtepec) to which it belongs.
The map also includes important towns of the region which were not sampled to be better orient the reader in space.

```{r map data, message = FALSE}
# read in map data
map_data <- read_tsv("https://raw.githubusercontent.com/SAuderset/MixteCaSo/main/data/metadata.tsv") %>%
  filter(MapAbbr!="abas")

# get min/max values for map box
summary(map_data$Latitude)
summary(map_data$Longitude)
```
```{r map setup, message = FALSE}
# set up map with stamen
mixtec_base <- get_stadiamap(bbox = c(left=-99.2, bottom=16, right=-96.55, top=18.75), maptype = "stamen_toner_background", zoom = 10, crop = TRUE)

# set up map with labels as function, to use as a base for plotting variables
map_all_labels <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(map_data, is.na(DOCULECT)), size = 2.5, fontface = "bold") +
  geom_label_repel(aes(x = Longitude, y = Latitude, label = MapAbbr), data = filter(map_data, !is.na(DOCULECT)), size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, color = "white", bg = "black", alpha = 0.5) +
  guides(color="none") +
  theme_map()
```
Next, I will set up a custom color scheme for the subgroups. For this, I first create a new column in the metadata with subgroups that shows lower-levels only for the very large Group 7.
```{r colors custom}
metadata <- read_tsv("https://raw.githubusercontent.com/SAuderset/MixteCaSo/main/data/metadata.tsv") %>%
  mutate(AudersetGroupC = case_when(AudersetGroup=="Group 7" & !is.na(AudersetGroupSub) ~ AudersetGroupSub,
                                    TRUE ~ AudersetGroup), .before = AudersetGroup) %>%
  mutate(AudersetGroupC = factor(AudersetGroupC, levels = c("Unclear", "Group 1", "Group 2", "Group 3", "Group 4", "Linkage 5", "Group 6", "Group 7", "Group 71", "Group 72", "Group 73", "Group 74", "Linkage 75", "Group 76")))

# custom color scale based on viridis palettes for 14 levels
ccol_14 <- c(
  "#6D4C3D",
  "#90d743",
  "#2B4570",
  "#35b779",
  "#43BCCD",
  "#4f772d",
  "#F9C80E",
  "#6e0930",
  "#e36596",
  "#bc3908",
  "#832161",
  "#c9071b",
  "#ba63ce", 
  "#f26419"
 )
```

We also set up a custom function to count number of changes easily and a custom template for the maps, using a unicode font.
```{r custom functions}
# function to calculate number of changes, number of NA, and ratio per variety
changes_calc <- function(c){
  c %>%
  group_by(DOCULECT) %>%
  mutate(n_change = sum(Value, na.rm = TRUE)) %>%
  mutate(n_na = sum(is.na(Value))) %>%
  mutate(r_change = round((100/(length(unique(Variable))-n_na))*n_change)) %>%
  ungroup() %>%
  distinct(DOCULECT, .keep_all = TRUE) %>%
  select(-Value) %>%
  mutate(Variable = str_match(Variable, "[[:upper:]]+")) %>%
  relocate(n_change:r_change, .after = DOCULECT)
}

# set up ggplot theme with unicode font and appropriate font size for reuse
theme_scmap <- theme(plot.title=element_markdown(family = "Noto Sans", size = 11, color = "white", fill = "black", linewidth = 5, margin = margin(t=40, b=-30, l=50), padding = unit(c(4, 4, 4, 4), "pt")),
        legend.title = element_blank(),
        legend.text = element_text(family = "Noto Sans", size = 10),
        legend.position = c(0, 0.5))
```

Now we can display a map of the sample colored by subgroups.
```{r map with colors}
# map with subgroups
map_subgroups <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(metadata, is.na(AudersetGroupC)), size = 2.5, fontface = "bold") +
  geom_label_repel(aes(x = Longitude, y = Latitude, label = MapAbbr, fill = AudersetGroupC), data = filter(metadata, !is.na(AudersetGroupC)), size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, alpha = 0.85) +
  scale_fill_manual(values = ccol_14) +
  guides(fill = guide_legend(title = "Subgroup", override.aes = aes(label = "", alpha = 1))) +
  theme_map() +
  theme(legend.title=element_text(size=9),
        legend.text=element_text(size=9),
        legend.key.size = unit(1,"line"),
        legend.position = c(0, 1),
        legend.justification = c(0, 1),
        legend.box.background = element_rect(color = "white"))
map_subgroups
```
We create a map of Mesoamerica with the sample map highlighted as an inset.
```{r mesoamerica inset}
# mesoamerica map
mesoamerica_map <- get_stadiamap(bbox = c(left = -112, bottom = 7.7, right = -82, top = 30), zoom = 5, maptype = "stamen_toner_lite")
# make square for where the inset is
square <- data.frame(
  lon = c(-99.2, -99.2, -96.2, -96.2, -99.2),  # Adjust these values for your square
  lat = c(16, 18.75, 18.75, 16, 16)  # Adjust these values for your square
)

# map with square
mesoamerica_inset <- ggmap(mesoamerica_map) +
  geom_polygon(data = square, aes(x = lon, y = lat), fill = NA, color = "black", size = 1) +
  theme_map()
mesoamerica_inset
```

Now we can combine the two maps.
```{r combine maps}
map_subgroups_inset = ggdraw() +
  draw_plot(map_subgroups) +
  draw_plot(mesoamerica_inset, x = 0.65, y = 0.01, width = 0.3, height = 0.3, valign = 0, halign = 0)
map_subgroups_inset
```

```{r test map, include=FALSE}
ggsave("map-subgroups-labels.png", map_subgroups_inset, device = "png", width = 21, units = "cm", dpi = 600, path = "/plots")
```


# Overview numbers
I generate some overview numbers for the paper regarding number of cognate sets, proto-forms, sound changes etc.

```{r numbers}
# number of reconstructed proto-forms, number of new reconstructions, number of classifier morphemes
# add variable that encodes whether the form was reconstructed by a previous source
protoforms <- read_tsv("https://raw.githubusercontent.com/SAuderset/MixteCaSo/main/data/protoforms.tsv") %>%
  mutate(previous_source = if_all(ends_with("ID"), ~is.na(.)))

# lexical entries in cognate sets
# set up vector with sample languages
sample_languages <- unique(var_seg_all$DOCULECT)
cognate_db <- read_tsv("https://raw.githubusercontent.com/SAuderset/mixtecan-cognate-database/main/mixtecan_cognate_db.tsv") %>%
  filter(DOCULECT %in% sample_languages)

# combine
numbers_overview <- tibble(
  pf_total = nrow(protoforms),
  pf_new = nrow(filter(protoforms, previous_source==TRUE)),
  pf_clf = nrow(filter(protoforms, str_detect(MEANING, "CLF"))),
  cognates = nrow(cognate_db),
  sc_total = ncol(var_seg_all),
  sc_vowels = ncol(select(var_seg_all, matches("A|E|I|O|U|Y"))),
  sc_cons = ncol(var_seg_all)-ncol(select(var_seg_all, matches("A|E|I|O|U|Y"))),
  sc_excluded = length(low_coverage_var)
)
numbers_overview
```


# Use cases

For the analysis and plotting, we create a data set that combines the change variables, metadata, and definitions.
```{r file comb}
# definition file
changes_def <- read_tsv("https://raw.githubusercontent.com/SAuderset/MixteCaSo/main/definitions/changes_segments.tsv")

# combine, recode to numeric
changes_combined <- var_seg %>%
  pivot_longer(-DOCULECT, names_to = "ID", values_to = "Value") %>% 
  left_join(., select(changes_def, ID:ENVIRONMENT_RIGHT)) %>%
  left_join(., select(metadata, DOCULECT, MapAbbr:AudersetGroupC, Latitude:Longitude)) %>%
  mutate(Value = case_when(Value=="yes" ~ 1,
                           Value=="no" ~ 0,
                           TRUE ~ as.numeric(Value)))
glimpse(changes_combined)
```
We also set up two custom functions for map plotting.
```{r map part func}
# custom map function for displaying variables
map.part.variable <- function(p, v){
  ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(metadata, is.na(AudersetGroupC)), size = 2.5, fontface = "bold") +
  geom_label_repel(data = p, aes(x = Longitude, y = Latitude, label = MapAbbr, fill = v), size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, color = "white")
}
# custom map theme
map.sc.theme <- function(){
  theme_map() +
  theme(legend.title=element_blank(), 
        legend.text=element_text(size=9),
        legend.key.size = unit(1,"line"),
        legend.position = c(0, 0.6))
}
```


## Summarizing reflexes of a proto-sound and conditioning environments
As an example for how reflexes and conditioning environments of a proto-sound can be summarized, we look at the developments from proto-Mixtec \*s. 
We subset our data set to just the changes from \*s and look at the distinct reflexes. We exclude the special cases of loss and metathesis, as we are interested in the primary reflexes. Retention is not coded as such in the database, because that does not constitute a change. However, the varieties that retain \*s are those that do not have any changes from \*s. We extract those from the database and create a new variable summarizing the main reflexes and retention.
We plot this on a map.

```{r summary refl}
# subset to s, keep only those that apply
pm_s <- changes_combined %>%
  filter(str_detect(ID, "S")) %>%
  select(DOCULECT:AudersetGroupC, SOUND_FROM:ENVIRONMENT_RIGHT, Latitude:Longitude)
# overview of modern reflexes
pm_s %>%
  distinct(SOUND_TO)
  
# recode for global changes excluding metathesis and initial loss
# get retention first
pm_s_retention <- pm_s %>%
  filter(!is.na(Value)) %>%
  group_by(DOCULECT) %>%
  mutate(s_reflexes = sum(Value)) %>%
  distinct(DOCULECT, s_reflexes, .keep_all = TRUE) %>%
  filter(s_reflexes==0) %>%
  mutate(s_reflexes = if_else(s_reflexes==0, "no change", "x"))
  
# changes
pm_s_global <- pm_s %>%
  filter(SOUND_FROM!="*si") %>%
  filter(SOUND_TO!="ø") %>%
  group_by(DOCULECT) %>%
  mutate(s_total = sum(Value)) %>%
  filter(Value==1) %>% 
  mutate(s_reflexes = paste(SOUND_TO, "/", ENVIRONMENT_LEFT, "_", ENVIRONMENT_RIGHT)) %>%
  mutate(s_reflexes = str_remove_all(s_reflexes, "NA")) %>%
  distinct(DOCULECT, s_reflexes, .keep_all = TRUE) %>%
  bind_rows(pm_s_retention) %>%
  group_by(DOCULECT) %>%
  mutate(s_reflexes = paste(s_reflexes, collapse = ", ")) %>%
  distinct(DOCULECT, s_reflexes, .keep_all = TRUE) %>%
  mutate(s_reflexes_short = case_when(
    s_reflexes=="ʃ /  _ i,e,ĩ,ẽ, ʃ / σ _ {i,ĩ}(ʔ)#" ~ "ʃ/_i,e",
    s_reflexes=="ð /  _ ɨ,e,a,o,u,ɨ̃,ẽ,ũ, ð /  _ i,ĩ" ~ "ð",
    s_reflexes=="ð /  _ ɨ,e,a,o,u,ɨ̃,ẽ,ũ, ʃ /  _ i,e,ĩ,ẽ, ʃ / σ _ {i,ĩ}(ʔ)#" ~ "ð/_ɨ, e,a,o,u | ʃ/_i",
    s_reflexes=="h /  _ ɨ,a,o,ɨ̃, h /  _ i,e,ĩ,ẽ, h / # _ u,ũ" ~ "h",
    s_reflexes=="h / # _ u,ũ, ʃ /  _ i,e,ĩ,ẽ, ʃ / σ _ {i,ĩ}(ʔ)#" ~ "h/_u, ʃ/_i,e",
    s_reflexes=="h /  _ ɨ,a,o,ɨ̃, h / # _ u,ũ, ʃ /  _ i,e,ĩ,ẽ, ʃ / σ _ {i,ĩ}(ʔ)#" ~ "h/_ɨ,a,o,u, ʃ/_i,e",
    TRUE ~ s_reflexes)) %>%
  ungroup() %>%
  mutate(s_reflexes_short = fct_relevel(s_reflexes_short, "no change", after = Inf)) %>%
  arrange(s_reflexes_short)
glimpse(pm_s_global)

# map showing distribution of global reflexes
s_refl_map <- map.part.variable(pm_s_global, pm_s_global$s_reflexes_short) +
  scale_fill_viridis(discrete = TRUE, option = "turbo", begin = 0.1, end = 0.9) +
  guides(fill = guide_legend(override.aes = aes(label = ""))) +
  map.sc.theme()
s_refl_map

# look at reflexes per group
s_refl_group <- pm_s_global %>%
  group_by(AudersetGroupC, s_reflexes) %>%
  summarize()
```
```{r s refl map save, include=FALSE}
ggsave("map-s-refl.png", s_refl_map, device = "png", height = 14, width = 17, units = "cm", dpi = 600, path = "/plots")
```


## Looking at specific types of changes
The database can also be used to look at specific types of changes, such as the loss of a sound or palatalization of different types of stops. We illustrate this here with the palatalization of proto-Mixtec *t before i.
This is coded as one change, so we subset the database to just that. Because we are interested in the distribution of this change across subgroups, we recode it as present/absent and use color to represent subgroups.
```{r t pal}
# subset
t_i_values <- changes_combined %>%
  filter(ID=="T01") %>%
  mutate(Value = as_factor(case_when(Value==0 ~ "absent",
                                     Value==1 ~ "present",
                                     TRUE ~ "NA"))) %>%
  mutate(Value = fct_relevel(Value, "present", "absent", "NA"))

# plot on map with subgroups
t_i_map <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(map_data, is.na(DOCULECT)), size = 2.5, fontface = "bold") +
  geom_point(aes(x = Longitude, y = Latitude, color = AudersetGroupC, shape = Value), data = t_i_values, size = 4) +
  scale_color_manual(values = ccol_14) +
  guides(shape = guide_legend(title = "*t > tʃ /_i"), color = guide_legend(title = "Subgroup", override.aes = list(size = 3), direction = "vertical", legend.position = "right")) +
  theme_map() +
  theme(legend.title=element_text(size=9), 
        legend.text=element_text(size=8),
        #legend.position = c(0, 1),
        #legend.justification = c("left", "top")
        )
t_i_map
```

```{r i refl map save, include=FALSE}
ggsave("map-t-pal-i.png", t_i_map, device = "png", height = 16, width = 21, units = "cm", dpi = 600, path = "/plots")
```


## Summarizing changes across varieties
We count the total number of sound changes per variety to provide an overview of which varieties are particularly conservative and particularly innovative. We do the same for consonants and vowels separately to check for possible differences between these two sound classes.
We do this by aggregating over the variables file with the numeric values (0 and 1).
```{r overview}
# merge variable file (numeric) with metadata
metadata_sub <- select(metadata, DOCULECT, MapAbbr:JosserandAreaSub, Latitude, Longitude)
# create numeric version of variables
var_num <- var_seg %>%
  mutate(across(matches("^[[:upper:]]{1}\\d{2}$"), ~case_match(.x, 
                                                               "yes" ~ 1,
                                                               "no" ~ 0))) %>%
  left_join(., metadata_sub)

# count overall number of changes per variety; add percentage taking into account the NAs
total_changes <- ncol(var_seg[, -1])
changes_total <- var_num %>%
  rowwise() %>%
  mutate(NumAll = sum(across(matches("^[[:upper:]]{1}\\d{2}$")), na.rm = TRUE)) %>%
  mutate(NAAll = sum(is.na(across(matches("^[[:upper:]]{1}\\d{2}$"))))) %>%
  mutate(PropAll = round(100/(total_changes-NAAll)*NumAll, 1)) %>%
  mutate(NumVow = rowSums(across(matches("^[AEIOUY]{1}\\d{2}$")), na.rm = TRUE)) %>%
  mutate(NAVow = sum(is.na(across(matches("^[AEIOUY]{1}\\d{2}$"))))) %>%
  mutate(PropVow = round(100/(total_changes-NAVow)*NumVow, 1)) %>%
  mutate(NumCons = rowSums(across(matches("^[DJKNSTWX]{1}\\d{2}$")), na.rm = TRUE)) %>%
  mutate(NACons = sum(is.na(across(matches("^[DJKNSTWX]{1}\\d{2}$"))))) %>%
  mutate(PropCons = round(100/(total_changes-NACons)*NumCons, 1)) %>%
  select(DOCULECT, NumAll:PropCons, MapAbbr:Longitude) %>%
  ungroup() %>%
  arrange(desc(PropAll))

# summary overview of range
changes_total %>% 
  select(NumAll:PropCons) %>% 
  summary()

# table of overall changes per variety
changes_total_table <- changes_total %>% 
  select(DOCULECT, matches("Num|Prop")) %>% 
  arrange(PropAll) %>%
  kable(., booktabs = TRUE, longtable = TRUE, escape = FALSE) %>%
  kable_styling(full_width = FALSE, font_size = 11)
changes_total_table
```

```{r plot all changes}
# map with proportion of changes
map_changes_overall <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(metadata, is.na(AudersetGroupC)), size = 2.5, fontface = "bold") +
  geom_label_repel(aes(x = Longitude, y = Latitude, label = MapAbbr, fill = PropAll), data = changes_total, size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, color = "white") +
  scale_fill_viridis(option = "inferno", end = 0.9, direction = -1) +
  theme_map() +
  theme(legend.title=element_blank(), 
        legend.text=element_text(size=9),
        legend.key.size = unit(1,"line"),
        legend.position = c(0, 0.7))
map_changes_overall

# map with total number of changes in consonants
map_changes_cons <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(metadata, is.na(AudersetGroupC)), size = 2.5, fontface = "bold") +
  geom_label_repel(aes(x = Longitude, y = Latitude, label = MapAbbr, fill = PropCons), data = changes_total, size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, color = "white") +
  scale_fill_viridis(option = "inferno", end = 0.9, direction = -1) +
  theme_map() +
  theme(legend.title=element_blank(), 
        legend.text=element_text(size=9),
        legend.key.size = unit(1,"line"),
        legend.position = c(0,0.7))
map_changes_cons

# map with total number of changes in vowels
map_changes_vow <- ggmap(mixtec_base) +
  geom_text(aes(x = Longitude, y = Latitude, label = Abbreviation), data = filter(metadata, is.na(AudersetGroupC)), size = 2.5, fontface = "bold") +
  geom_label_repel(aes(x = Longitude, y = Latitude, label = MapAbbr, fill = PropVow), data = changes_total, size = 3.5, fontface = "bold", family = "Linux Libertine", max.overlaps = 40, box.padding = 0, point.padding = 0.1, label.padding = 0.1, color = "white") +
  scale_fill_viridis(option = "inferno", end = 0.9, direction = -1) +
  theme_map() +
  theme(legend.title=element_blank(), 
        legend.text=element_text(size=9),
        legend.key.size = unit(1,"line"),
        legend.position = c(0,0.7))
map_changes_vow
```
```{r change map, include=FALSE}
ggsave("map-overview-all.png", map_changes_overall, device = "png", height = 14, width = 17, units = "cm", dpi = 600, path = "/plots")
ggsave("map-overview-cons.png", map_changes_cons, device = "png", height = 14, width = 17, units = "cm", dpi = 600, path = "/plots")
ggsave("map-overview-vowels.png", map_changes_vow, device = "png", height = 14, width = 17, units = "cm", dpi = 600, path = "/plots")
```


# References

Kahle, David \& Wickham, Hadley (2013). ggmap: Spatial Visualization with ggplot2. The R Journal 5(1), pp. 144–161. https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
